High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths

### 2021-10-11 manuscript/Manuscript.tex
manuscript update - need to finish the last discussion section and conclusion

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}

\linenumbers

@@

Earth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. This self-sustaining geodynamo is collectively powered by the heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. Paleomagnetic studies have found that a dynamo may have existed by the Archean \cite{Tarduno2014a}, and possibly as far back as the Hadean \cite{Tarduno2020a}. However, some estimates of the thermal conductivity of the core suggest that the nucleation of the inner core could be younger than 1 Ga, more recent than previous thought \cite{Pozzo2012a, Gomi2013a, Labrosse2015a, Ohta2016a}. Without the power of compositional convection provided by the process of light elements being rejected into the outer core from the crystallizing inner core, the geodynamo was likely dominantly driven by thermal conduction through the CMB, with perhaps additional stirring by other power sources such Mg precipitation \cite{ORourcke2016a}, or perhaps there were intervals early in Earth history when the dynamo was generated by exotic processes such as a convecting basal magma ocean \cite{Ziegler2013a}. Long-term secular cooling of Earth has the potential to have decreased the power available to the geodynamo through time prior to the nucleation of the inner core \cite<e.g>{Davies2021a}. As a result, it is possible that strength of the geomagnetic axial dipole field would have experienced a long-term decay before the emergence of the solid inner core after which it would have strengthened \cite{Aubert2009a, Davies2021a}. However, the estimates for thermal conductivity values for the core continue to vary\cite{Konopkova2016a, Ohta2016a}. Obtaining an observational record of the timing of inner core formation would constrain the thermal evolution of the core and mantle, in addition to improving our understanding of the history of Earth’s dynamo.

Paleomagnetic data from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Paleointensity studies have found very low axial dipole moment (VDM) estimates from rocks formed in the late Ediacaran \cite{Bono2019a, Shcherbakova2019a, Thallner2021b}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Other studies focusing on paleomagnetic directions have also found that the late Ediacaran might have been a period of time when the geomagnetic field were experiencing frequent reversals and waning of dipole dominance \cite{Bono2015a, Kodama2020a}. It has been interpreted that these findings are consistent with the results of geodynamo models using high core thermal conductivity values that suggest the inner core nucleation age being near the Ediacaran-Cambrian boundary \cite{Driscoll2016a, Davies2021a}. \citeA{Bono2019a} based on a polynomial fit through filtered Precambrian paleointensity database (PINT; \url{http://earth.liv.ac.uk/pint/}) to suggest the selected observational records show that the geodynamo power progressively decreased through the Proterozoic. They also recovered very low paleointensity values from \textit{ca.} 565 Ma single silicate crystals and interpreted them to be reflecting that the geodynamo reached its lowest near the Ediacaran-Cambrian boundary. Such decay trend in the Neoproterozoic is recently supported by new low paleointensity estimates derived from the \textit{ca.} 720 Ma Franklin Large Igneous Province \cite{Lloyd2021a}.

However, previous paleointensity data from intrusive and extrusive rock units in the Mesoproterozoic Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record Earth's virtual axial dipole moment (VADM) values similar to that of today, which are higher than the prediction of the long term decay curve (Fig. \ref{}). \citeA{Pesonen1983b} obtained an average virtual dipole moment of 93 ZAm$^2$ after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the Maimainse Point volcanics, intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flow of the \textit{ca.} 1096 Ma North Shore Volcanic Group. \citeA{Kulakov2013a} obtained an average VADM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by \citeA{Pesonen1983b, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \citeA{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas \citeA{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \citeA{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VADM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}. This high VADM value estimate from the Midcontinent Rift volcanics may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

Although it is typical of paleointensity experiments to suffer from low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested plagioclase crystals can protect magnetic minerals that exsolved within them against thermochemical alteration during paleointensity experiments. In this study, we present new paleointensity results from ``IZZI" absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the Beaver River anorthosite xenoliths has exceptionally high success rate. Thermal remanent magnetization (TRM) anisotropy and acquisition experiments reveal that the magnetite-dominated remanence of the anorthosite specimens has low TRM anisotropy and linear TRM acquisition. Magnetic imaging shows that the anorthosite specimens to have and dominant magnetic carriers interstitial to the plagioclase and are well shielded. With more single-slope, high quality paleointensity data, we seek to compare paleointensity records from the Midcontinent Rift intrusives with those from the extrusives; to test the hypothesis that the Earth's geomagnetic dipole strength remained at high values throughout the Midcontinent Rift; and to further enrich the reliable paleointensity records during the $\sim$25 myr of Midcontinent Rift magmatic activity.

@@

The Mesoproterozoic North American Midcontinent Rift (MCR) is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; \citeA{Vervoort2007a, Miller2013a}) was punctuated with rapid and voluminous emplacement of intrusive rocks as well as extrusive volcanics. Based on high-precision $^{206}$Pb/$^{238}$U geochronology and paleomagnetic data, \citeA{Swanson-Hysell2021a} hypothesized that the massive Duluth Complex and the coeval North Shore Volcanic group (Figure \ref{fig:Geologic_map}) formed as a result of having upwelling mantled-derived plume material encountering Laurentia's thinned lithosphere due to rifting, following an ``upside-down” drainage" which rapidly transported melt to today's North Eastern Minnesota (Figure \ref{fig:Geologic_map}; \citeA{Sleep1997a, Swanson-Hysell2014a}). The plagioclase-phyric Anorthositic Series rocks of the Duluth Complex that outcrop along an arcuate shape along Northeastern Minnesota (Figure \ref{fig:Geologic_map}) are a result of such vigorous magmatic activity.

@@

The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex punctuates another period of Midcontinent Rift main stage magmatism. $^{206}$Pb/$^{238}$U zircon dates and cross-cutting relationships tightly constraints the areally extensive Beaver River diabase of the Beaver Bay Complex to have been emplaced at 1091.7 $\pm$ 0.2 Ma \cite{Zhang2021b}. The Beaver River diabase intruded into hypabyssal depths via wide conduits that accommodated the entrainment of numerous anorthosite xenoliths that have short-axis diameters up to 180 meters. Geochronological, paleomagnetic, geochemical, and petrographic data are consistent with the hypothesis that the Beaver River diabase magma extruded to the surface, feeding the massive Greenstone Flow - one of the largest lava flows on Earth \cite{Doyle2016a, Zhang2021b}.

@@

Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. Moreover, the anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Fig. \ref{Petro_QDM}), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texture characterized by equigranular plagioclase crystals (Fig, \ref{Petro_QDM}). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Vermicular titanomagnetite and ilmenite symplectic intergrowths also exist surrounding accessory olivine or pyroxene in the Duluth Complex Anorthositic Series rocks (Fig. \ref{Petro_QDM}).

\begin{figure}
\noindent\includegraphics[width=\textwidth]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from Duluth Complex Anorthositic Series (left column) and Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (right column; \citeA{Zhang2021b}). Petrographic images (A-D) show Duluth Complex anorthosite have euhedral, elongate plagioclase crystals with interlocking texture and strong igneous foliation. The plagioclase can contain abundant Fe-Ti oxide needles (C) that have preferred orientations that are often parallel with the (010) planes of plagioclase. In contrast, plagioclase of sample MS99033 do not enclose oxide needles. Magnetic field maps acquired by the quantum diamond microscope (QDM) show that the oxide needles within Duluth plagioclase are strongly magnetic as evidenced through the overlapping magnetic dipole fields exerted above the oxides in (E). In comparison, the plagioclase of MS99033 is not magnetic. }}
\label{fig:Petro_QDM}
\end{figure}

Similar occurrences of magnetite and ilmenite inclusions within plagioclase have been described by \citeA{Haggerty1976a} in slowly cooled, mafic intrusions, and \cite{Scofield1986a} hypothesized that the inclusions acted as a means of buffering the melt with respect to oxygen, forming either crystallites along preexisting plagioclase substrate or as iron incorporated into the plagioclase that subsequently exsolved at subsolidus temperatures. Such a model requires that plagioclase, instead of other mafic minerals such as olivine, to be the first to crystallize from a magma and act as a sink for iron. The two-stage formation of the Duluth Complex Anorthositic Series fits this model \cite{Miller1990a}. During the first stage, high pressure at subcrustal depths (15-25 km depth) caused fractionation of Duluth Complex magma resulting in the formation of floating plagioclase mushes \cite{Kushiro1980a}. The second stage of crystallization then occurs at low pressures after the magma intruded into shallower chambers.

This proposed magmatic evolution of the Duluth Complex is in distinct contrast to the emplacement of the Beaver River diabase which rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \citeA{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported into shallow magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Fig. \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite interstitial to the plagioclase are typical of Beaver River anorthosites (Fig, \ref{}).

@@

We collected paleomagnetic cores that are 2.5 cm in diameter along the southern and eastern Beaver Bay Complex with a particular focus on acquiring paired sites of anorthosite xenoliths and their local diabase hosts during summer field seasons in 2019 and 2020. Sample cores were collected using a hand-held gasoline-powered drill and were oriented using a magnetic compass as well as a sun compass when possible. Sun compass orientations were preferentially used for determining the sample azimuth. Specimens from a total of 7 diabase and 14 anorthosite sites were used for paleointensity experiments (Table \ref{tab:Pmag_site_data}). Sister specimens underwent step-wise alternating field (AF) or thermal demagnetization at the UC Berkeley Paleomagnetism Lab and the results are presented in \citeA{Zhang2021b}. Paleomagnetic direction measurement level data are available within the MagIC database (\url{https://earthref.org/MagIC/doi/10.1029/2021GC009909}). The close directional similarities between the anorthosite xenolith and their host diabase, the overall cooling unit mean directions between the two lithologies, and thermal modeling results are consistent with the interpretation that the Beaver River diabase and anorthosite xenoliths acquired primary thermal remanent magnetization \textit{ca.} 1092 Ma during cooling after the emplacement of the diabase \cite{Zhang2021b}. Based on the anorthosite thermal demagnetization results we selected sites whose unblocking temperature ranges are narrow and near 580\textdegree C. Beaver River diabase sites with minimal secondary remanence component were also selected for paleointensity experiments.

\subsection*{Paleointensity experiment}
A total of 69 specimens from 7 diabase sites and a total of 86 specimens from 14 anorthosite xenoliths underwent paleointensity experiments that followed the step-wise double-heating Thellier method \cite{Thellier1959a} using the IZZI protocol \cite{Tauxe2004a} with heating steps up to 585 \textdegree C. Partial thermal remanent magnetization (pTRM) checks were performed systematically throughout the experiment to test whether there was significant mineralogical alteration due to heating and were assessed using the SCAT parameter of \citeA{Shaar2013a}. On top of the IZZI modification of the Thellier experimental protocol, we also performed a comparative study where we added an extra step of 20 mT alternating field (AF) cleaning on some of the specimens after each in-field step. The purpose is to study whether the AF cleaning could help improve experiment success rate by removing the remanence component carried by materials such as multi-domain (MD) or vortex state grains that often contributes to non-ideal paleointensity behaviors. All remanence measurements were made on a 2G Enterprises DC-SQUID superconducting rock magnetometer equipped with an automated sample changer system at the UC Berkeley Paleomagnetism laboratory. The magnetometer is housed inside a three-layer magnetostatic shield that maintains background fields of less than 500 nT. Heating steps were performed using an ASC TD-48SC thermal demagnetizer with a controlled field coil that allows for a magnetic field to be generated in the oven in conjunction with a DC power supply. The thermal demagnetizer was degaussed with an alternating field in the axial orientation following each in-field step such that residual fields within the oven were $<$10 nT during zero-field steps. Samples were placed in the same location within the thermal demagnetizer for each heating step and were maintained in the same orientation with regard to the applied field. During each heating step, samples remained at peak temperatures for 20 min. An applied laboratory field of 30 μT was used for all in-field steps. All heating steps were performed in air. The temperature increments for the experiments were chosen to cover characteristic remanent magnetizations held by (titano)magnetite informed by the previous demagnetization data, with smaller increment temperature steps performed close to the expected unblocking temperature of stoichiometric magnetite \cite{Zhang2021b}.

After the paleointensity experiment were finished, the same specimens underwent additional TRM anisotropy check with full TRMs given at 0, 30, 60, 90, 120, 150 degrees in the horizontal plane in specimen coordinates. After that, full TRM acquisition experiments were performed on the same specimens with applied fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axis.

>>>>

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\usepackage{adjustbox}
\linenumbers

@@

Earth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. This self-sustaining geodynamo is collectively powered by the heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. Paleomagnetic studies have found that a dynamo may have existed by the Archean \cite{Selkin2007a, Tarduno2014a, Brenner2020a}. However, some estimates of the thermal conductivity of the core suggest that the nucleation of the inner core could be younger than 1 Ga, more recent than previous thought \cite{Pozzo2012a, Gomi2013a, Labrosse2015a, Ohta2016a}. Without the power of compositional convection provided by the process of light elements being rejected into the outer core from the crystallizing inner core, the geodynamo was likely dominantly driven by thermal conduction through the CMB, with perhaps additional stirring by other power sources such Mg precipitation \cite{ORourcke2016a}, or perhaps there were intervals early in Earth history when the dynamo was generated by exotic processes such as a convecting basal magma ocean \cite{Ziegler2013a}. Long-term secular cooling of Earth has the potential to have decreased the power available to the geodynamo through time prior to the nucleation of the inner core \cite<e.g>{Davies2021a}. As a result, it is possible that rigor of the geomagnetic axial dipole field would have experienced a long-term decay before the emergence of the solid inner core after which it would have strengthened \cite{Aubert2009a, Davies2021a}. However, the estimates for thermal conductivity values for the core continue to vary\cite{Konopkova2016a, Ohta2016a}. Obtaining an observational record of the timing of inner core formation would constrain the thermal evolution of the core and mantle, in addition to improving our understanding of the history of Earth’s dynamo.

Paleomagnetic data from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Some Precambrian absolute paleointensity records from the PINT database (\url{http://earth.liv.ac.uk/pint/}) show evidence of an existence of an appreciable geomagnetic field prior to the Ediacaran. On the other hand, studies have found very low axial dipole moment (VDM) estimates from rocks formed in the late Ediacaran \cite{Bono2019a, Shcherbakova2019a, Thallner2021b}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Other studies focusing on paleomagnetic directions have also found that the late Ediacaran might have been a period of time when the geomagnetic field were experiencing frequent reversals and waning of dipole dominance \cite{Bono2015a, Kodama2020a}. It has been interpreted that these findings are consistent with the results of geodynamo models using high core thermal conductivity values that suggest the inner core nucleation age being near the Ediacaran-Cambrian boundary \cite{Driscoll2016a, Davies2021a}. \citeA{Bono2019a} based on a polynomial fit through filtered Precambrian paleointensity database to suggest the selected observational records show that the geodynamo power progressively decreased through the Proterozoic (Figure \ref{fig:MCR_VADM}). That study also recovered very low paleointensity values from \textit{ca.} 565 Ma single silicate crystals and interpreted them to be reflecting that the geodynamo reached its lowest near the Ediacaran-Cambrian boundary. New low paleointensity estimates derived from the \textit{ca.} 720 Ma Franklin Large Igneous Province are interpreted to support the existence of such decay trend in the Neoproterozoic \cite{Lloyd2021a}.

However, previous paleointensity data from intrusive and extrusive rock units in the Mesoproterozoic Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record Earth's virtual axial dipole moment (VADM) values similar to that of today, which are higher than the prediction of the long term decay curve (Figure \ref{fig:MCR_VADM}). \citeA{Pesonen1983b} obtained an average virtual dipole moment of 93 ZAm$^2$ after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the Maimainse Point volcanics, intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flow of the \textit{ca.} 1096 Ma North Shore Volcanic Group. \citeA{Kulakov2013a} obtained an average VADM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by \citeA{Pesonen1983b, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \citeA{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas \citeA{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \citeA{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VADM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}. This high VADM value estimate from the Midcontinent Rift volcanics may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

Although it is typical of paleointensity experiments to suffer from low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested that plagioclase crystals can protect magnetic minerals that exsolved within them against post-formation weathering and thermochemical alteration during paleointensity experiments \cite{Tarduno2005a}. In this study, we present new paleointensity results from ``IZZI" modified absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the Beaver River anorthosite xenoliths have a very high success rate. The consistent specimen- and site-level paleointensity results reveal that the magnetite-dominated remanence of the anorthosite specimens has low thermal remanent magnetization (TRM) anisotropy and linear TRM acquisition. Magnetic imaging shows that the anorthosite specimens to have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. With more single-slope, high quality paleointensity data, we seek to compare paleointensity records from the Midcontinent Rift intrusives with those from the extrusives; to test the hypothesis that the Earth's geomagnetic dipole strength remained at high values throughout the Midcontinent Rift; and to further enrich the reliable paleointensity records during the $\sim$25 myr of Midcontinent Rift magmatic activity.

@@

The Mesoproterozoic North American Midcontinent Rift (MCR) is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; \citeA{Vervoort2007a, Miller2013a}) was punctuated by rapid and voluminous emplacement of intrusive rocks as well as extrusive volcanics. Based on high-precision $^{206}$Pb/$^{238}$U geochronology and paleomagnetic data, \citeA{Swanson-Hysell2021a} hypothesized that the massive Duluth Complex and the coeval North Shore Volcanic group (Figure \ref{fig:Geologic_map}) formed as a result of having upwelling mantled-derived plume material encountering Laurentia's thinned lithosphere, following an ``upside-down” drainage" which rapidly funneled melt to today's North Eastern Minnesota (Figure \ref{fig:Geologic_map}; \citeA{Sleep1997a, Swanson-Hysell2014a}). The plagioclase-phyric Anorthositic Series rocks, the crystal-poor Layered Series intrusions of the Duluth Complex, and the coeval volcanics of the North Shore Volcanic Group (NSVG) that outcrop along Northeastern Minnesota (Figure \ref{fig:Geologic_map}) are interpreted to be a result of such vigorous magmatic activity.

@@

The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex punctuates another period of magmatism during the Midcontinent Rift main stage activity. $^{206}$Pb/$^{238}$U zircon dates and cross-cutting relationships tightly constraints the areally extensive Beaver River diabase of the Beaver Bay Complex to have been emplaced at 1091.7 $\pm$ 0.2 Ma \cite{Zhang2021b}. The Beaver River diabase intruded into hypabyssal depths via wide conduits that accommodated the entrainment of numerous anorthosite xenoliths that have short-axis diameters up to 180 meters \cite{Boerboom2004a, Boerboom2006b}. Geochronological, paleomagnetic, geochemical, and petrographic data are consistent with the hypothesis that the Beaver River diabase magma extruded to the surface, feeding the massive Greenstone Flow - one of the largest lava flows on Earth \cite{Doyle2016a, Zhang2021b}.

@@

Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. Moreover, the anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Figure \ref{Petro_QDM}A), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texture characterized by equigranular plagioclase crystals (Figure \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial vermicular titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from Duluth Complex Anorthositic Series (left column) and Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (right column; \citeA{Zhang2021b}). Petrographic images (A-D) show Duluth Complex anorthosite have euhedral, elongate plagioclase crystals with interlocking texture and igneous foliation. (C) The plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the (010) planes of plagioclase. In contrast, plagioclase of sample MS99033 do not enclose large oxide needles. Magnetic field maps acquired by the quantum diamond microscope (QDM) show that the oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic dipole fields exerted above the oxides in (E). In comparison, relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite. The dipole-like features in (F) do not show strong preferred orientations. Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
\end{figure}

Occurrences of magnetite and ilmenite symplectites and needles within anorthositic rocks have been described by \citeA{Scofield1975a, Scofield1986a} in the Wichita Mountains and by \citeA{Haggerty1976a} in other slowly cooled, mafic intrusions. \citeA{Scofield1986a} hypothesized that the inclusions acted as a means of buffering the melt with respect to oxygen, forming either crystallites along preexisting plagioclase substrate or as iron incorporated into the plagioclase that subsequently exsolved at subsolidus temperatures. Such a model requires that plagioclase, instead of other mafic minerals such as olivine, to be the first to crystallize from a magma and act as a sink for iron. The two-stage formation of the Duluth Complex Anorthositic Series fits this model \cite{Miller1990a}. During the first stage, high pressure at subcrustal depths (15-25 km depth) caused fractionation of Duluth Complex magma resulting in the formation of floating plagioclase mushes \cite{Kushiro1980a}. The second stage of crystallization then occurs at low pressures after the magma intruded into shallower chambers.

This proposed magmatic evolution of the Duluth Complex is in distinct contrast to the emplacement of the Beaver River diabase which rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \citeA{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported by upward currents into hypabyssal magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Figure \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Figure \ref{Petro_QDM}).

@@

We collected paleomagnetic cores that are 2.5 cm in diameter along the southern and eastern Beaver Bay Complex with a particular focus on acquiring paired sites of anorthosite xenoliths and their local diabase hosts during summer field seasons in 2019 and 2020. Sample cores were collected using a hand-held gasoline-powered drill and were oriented using a magnetic compass as well as a sun compass when possible. Sun compass orientations were preferentially used for determining the sample azimuth. Specimens from a total of 7 diabase and 14 anorthosite sites were used for paleointensity experiments (Table \ref{tab:Pmag_site_data}). Sister specimens underwent step-wise alternating field (AF) or thermal demagnetization at the UC Berkeley Paleomagnetism Lab and the results are presented in \citeA{Zhang2021b}. The close directional similarities between the anorthosite xenoliths and their host diabase and thermal modeling results are consistent with the interpretation that the Beaver River diabase and anorthosite xenoliths acquired primary thermal remanent magnetization \textit{ca.} 1092 Ma during cooling after the emplacement of the diabase \cite{Zhang2021b}. Based on the anorthosite thermal demagnetization results we selected sites whose unblocking temperature ranges are narrow and near 580\textdegree C for paleointensity experiments. Beaver River diabase sites with minimal secondary remanence were also selected for paleointensity experiments.

\subsection*{Paleointensity experiment}
A total of 86 specimens from 14 anorthosite xenoliths and a total of 69 specimens from 7 diabase sites underwent paleointensity experiments that followed the step-wise double-heating Thellier method \cite{Thellier1959a} using the IZZI protocol \cite{Tauxe2004a} with heating steps up to 585 \textdegree C. Partial thermal remanent magnetization (pTRM) checks were performed systematically throughout the experiment to test whether there was significant mineralogical alteration due to heating and were assessed using the SCAT parameter of \citeA{Shaar2013a}. On top of the IZZI modification of the Thellier experimental protocol, we also performed a comparative study where we added an extra step of 20 mT alternating field (AF) cleaning on some of the specimens after each in-field step. The purpose is to study whether the AF cleaning could help improve experiment success rate by removing the remanence component carried by materials such as multi-domain (MD) or vortex state grains that often contributes to non-ideal paleointensity behaviors. All remanence measurements were made on a 2G Enterprises DC-SQUID superconducting rock magnetometer equipped with an automated sample changer system at the UC Berkeley Paleomagnetism laboratory. The magnetometer is housed inside a three-layer magnetostatic shield that maintains background fields of less than 500 nT. Heating steps were performed using an ASC TD-48SC thermal demagnetizer with a controlled field coil that allows for a magnetic field to be generated in the oven in conjunction with a DC power supply. The thermal demagnetizer was degaussed with an alternating field in the axial orientation following each in-field step such that residual fields within the oven were $<$10 nT during zero-field steps. Samples were placed in the same location within the thermal demagnetizer for each heating step and were maintained in the same orientation with regard to the applied field. During each heating step, samples remained at peak temperatures for 20 min. An applied laboratory field of 30 $\mu$T was used for all in-field steps. All heating steps were performed in air. The temperature increments for the experiments were chosen to cover characteristic remanent magnetizations held by (titano)magnetite informed by the previous demagnetization data, with smaller increments performed close to the expected unblocking temperature of stoichiometric magnetite \cite{Zhang2021b}.

% After the paleointensity experiment were finished, the same specimens underwent additional TRM anisotropy checks with full TRMs given at 0, 30, 60, 90, 120, 150 degrees in the horizontal plane in specimen coordinates. After that, full TRM acquisition experiments were performed on the same specimens with applied fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axis.
